---
title: "Data Analysis and Statistical Inference Project Proposal"
author: "Arindam Biswas"
date: "today's date"
output:
  html_document:
    keep_md: yes
    theme: cerulean
  word_document: default
---
## RESEARCH QUESTION: 

### In one sentence, what is your research question? 

Is there a relationship between race of respondent and number of children in the household?

## DATA - Citation: 

### Include a citation for your data, and if your data set is online, provide a link to the source.

The data set to be used for this project is General Social Survey Cumulative File, 1972-2012 Coursera Extract, which is modified for Data Analysis and Statistical Inference course (Duke University). 

The data set can be downloaded from http://bit.ly/dasi_gss_data

**Data Citation**

Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012 [Cumulative File]. ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research, University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributors], 2013-09-11. doi:10.3886/ICPSR34802.v1

Persistent URL: http://doi.org/10.3886/ICPSR34802.v1

## DATA - Collection: 

### Describe how the data were collected.

The data has been collected by **NORC at the University of Chicago**. The data was collected through a sociological survey **General Social Survey (GSS)** used to collect data on demographic characteristics and attitudes of residents of the United States. The data has been collected cumulatively from 1972 to 2012. 

The data has a total of 57,061 cases and 114 variables. However the number of variables differ over the years as the number of variables in the initial years were fewer and over the years new questions has been added to the survey and hence new variables are added to the data set.

The codebook available at https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html lists all variables, the values they take, and the survey questions associated with them.

**Note**
Not all respondents answered all questions in all years and in this data set all missing values have been re-coded to missing (NA).

## DATA - Cases (observational/experimental units): 

### What are the cases? (Remember: case = units of observation or units of experiment)

The cases are adult resident of United States, who in their household have provided responses for the GSS survey questions pertaining to **race: Race of respondent** and **childs: Number of children** 

## DATA - Variables: 

### What are the two variables you will be studying to answer the research you stated earlier? State the type of each variable.

The data variables to be studied are:

- race: Race of respondent 
- childs: Number of children

**Type of variable**

- race: categorical, non-ordinal
- childs: numerical, discrete

## DATA - Type of study: 

### What is the type of study? Is it an observational study or an experiment? Explain how you've arrived at your conclusion using information on the sampling and/or experimental design.

As a sociological survey has been used to collect data on demographic characteristics and attitudes of residents of the United States. So the data was collected in a way that does not directly interfere with how the data arises, hence the study is observational. 

## DATA - Scope of inference - generalizability: 

### Identify the population of interest, and whether the findings from this analysis can be generalized to that population, or, if not, a subsection of that population. Explain why or why not. Also discuss any potential sources of bias that might prevent generalizability.

As the survey was designed to administered randomly to any adult resident of United States, that indicates that random sampling was used to administer the survey, hence the results can be generalized to the entire the population.

While there was potential source of bias arising from non-response to the survey. However broadly the non-response to the survey has been also random and hence we can establish that the survey was responded by a sample set of respondents who are representative of the population.

## DATA - Scope of inference - causality: 

###Can these data be used to establish causal links between the variables of interest? Explain why or why not.

As the study is observational, so at best we can only establish association between the variables of interest. Whereas if the study was an experiment, the data could have been used to establish causal links between the variables of interest.

## EXPLORATORY DATA ANALYSIS:

### Perform a brief exploratory data analysis - just one or two relevant descriptive statistics and visualizations of the data. Also address what the exploratory data analysis suggests about your research question. 

First we use the following code to load the GSS dataset into R
```{r echo=FALSE}
load(url("http://bit.ly/dasi_gss_data"))
```

Now we load the libraries required for the exploratory data analysis
```{r echo=FALSE}
library(ggplot2)
```

Then we study the list of the variable names in the dataset
```{r echo=FALSE}
names(gss)
```

Then we analyse the structure of the two variables of interest - `race` and `childs`. We observe that `race` is categorical and non-ordinal. While `childs` is numerical and discrete
```{r echo=FALSE}
str(gss$race)
str(gss$childs)
```

Then we create a subset of the data using dataframe gss_subset with only the two variables of interest - `race` and `childs` and set the column headers as Race and ChildCount
```{r echo=FALSE}
gss_subset <- data.frame(Race=gss$race, ChildCount=gss$childs)
```

Now we remove those rows from the data frame gss_subset where either `race` or `childs` have missing (NA) values
```{r echo=FALSE}
gss_subset <- gss_subset[!(is.na(gss_subset$Race)) & !(is.na(gss_subset$ChildCount)),]
```

Then we check the number of cases/observations in the gss_subset data frame, which is 56880
```{r echo=FALSE}
nrow(gss_subset)
```

Then we summarize the race(categorical variable) using a table and a bar plot. We observe that White race has maximum number of respondents while Other race has least.
```{r echo=FALSE}
table(gss_subset$Race)

plot(gss_subset$Race, main="Table of Race of Respondent Households", xlab="Race", ylab="Count of Respondent Households")
```

Then we summarize the childs(numerical variable) using a summary and box plot
```{r echo=FALSE}
summary(gss_subset$ChildCount)
boxplot(gss_subset$ChildCount, main = "Box Plot of No of Childs in Respondent Household",
        ylab = "No of Childs",col = "lightblue")
```

Now we analyze if there is a relationship between the two variables `race` and `childs` using boxplot
```{r echo=FALSE}
plot(gss_subset, main="Box Plot of No of Childs in Respondent Household by Race", xlab="Race", ylab="No of Childs", col="pink")
```

We observe that while median value of no of childs is similar across the three races, there is significant difference in the overall pattern and Inter Quartile Range (ICR) for no of childs across the three races. The IQRs for `White` and `Other` races are wider than the IQR for `Black` race. Also `White`and `Other` races have higher maximum number childs and fewer outliers as compared to `Black` race.

## Attach a page of your data set that includes all of the relevant columns. You do not need to include all of the data; one page is sufficient.

Now we list out the first 50 rows of the data frame gss_subset containing the columns `race` and `childs`
```{r echo=FALSE}
gss_subset2 <- data.frame(gss$race, gss$childs)
head(gss_subset2, 50)
```